Linearizability Proofs Based On Forward Simulations

### 2017-01-19 draftCAV/queue.tex
started stacks

Informally, $AbsQ$ records the happens-before order between enqueue operations for which the added value has not been removed by a dequeue operation. The linearization point of a dequeue operation with return value $d\neq{\tt EMPTY}$ is enabled only if the happens-before order stored in the current state contains a minimal enqueue operation that adds the value $d$. The effect of the linearization point is that the minimal enqueue is removed from the current state and the return value is recorded in the library state. When the return value is {\tt EMPTY}, the linearization point of a dequeue is enabled only if the current state stores only pending enqueue operation. The return of a dequeue is enabled only if the returned value matches the one fixed at the linearization point. 
%\textcolor{red}{Important: I think EMPTY return is problematic and I defined it wrong in my machine too. We should be able to return EMPTY if there are only pending nodes. Consider the following history of $AbsQ_0$: $inv(enq, d_1,k_1), inv(enq, d_2,k_2), inv(deq, k_3), lin(deq, \texttt{EMPTY}, k_3)$. This history should be reflected in $AbsQ$ by enabling lin \texttt{EMPTY} of dequeue when there are pending nodes. We also need to update the rules in figure.}

\begin{wrapfigure}{l}{7cm}
\vspace{-8mm}
\includegraphics[width=7cm]{fig-queue12.pdf}
%
%\vspace{2mm}
%\includegraphics[width=7cm]{fig-queue2.pdf}
\vspace{-5mm}
\caption{Simulating queue histories with $AbsQ$.}
\label{fig:queueSim}
\vspace{-10mm}
\end{wrapfigure}
Figure 2 pictures two executions of $AbsQ$ for two extended histories (that include dequeue linearization points). The state of $AbsQ$ after each action is pictured as a graph below the action. The nodes of this graph represent enqueue operations and the edges happens-before constraints. Each node is labeled by a value (the input of the enqueue) and a flag {\tt PEND} or {\tt COMP} showing whether the operation is pending or completed. For instance, in the case of the first history, the dequeue linearization point $lin(deq,y,3)$ is enabled because the current happens-before contains a \emph{minimal} enqueue operation with input $y$. Note that a linearization point $lin(deq,x,3)$ is also enabled at this state.

@@

We show that $\mathit{fs}$ is indeed a $C\cup R\cup Lin(deq)$-forward simulation. Let $s$ and $t$ be states of $\mathit{HWQ}$ and $AbsQ$, respectively, such that $(s,t)\in\mathit{fs}$. 
We omit discussing the trivial case of transitions labeled by call and return actions (for the return a dequeue operation $k$, we use the equality between the local variable ${\tt x}(k)$ in $s$ and the component $rv(k)$ in $t$). \textcolor{red}{ I think it is good to mention again that call/return actions in HWQ correspond to the same call/return actions in AbsQ (without any other internal action). I also think that invoke enqueue operation is non-trivial. Preservation of the strict partial order and all of the above items a, b and c needs to be rechecked.}

>>>>

Informally, $AbsQ$ records the happens-before order between enqueue operations for which the added value has not been removed by a dequeue operation. The linearization point of a dequeue operation with return value $d\neq{\tt EMPTY}$ is enabled only if the happens-before order stored in the current state contains a minimal enqueue operation that adds the value $d$. The effect of the linearization point is that the minimal enqueue is removed from the current state and the return value is recorded in the library state. When the return value is {\tt EMPTY}, the linearization point of a dequeue is enabled only if the current state stores only pending enqueue operations (the dequeue overlaps with all the enqueue operations stored in the current state and it can be linearized before all of them).  
The return of a dequeue is enabled only if the returned value matches the one fixed at the linearization point. 
%\textcolor{red}{Important: I think EMPTY return is problematic and I defined it wrong in my machine too. We should be able to return EMPTY if there are only pending nodes. Consider the following history of $AbsQ_0$: $inv(enq, d_1,k_1), inv(enq, d_2,k_2), inv(deq, k_3), lin(deq, \texttt{EMPTY}, k_3)$. This history should be reflected in $AbsQ$ by enabling lin \texttt{EMPTY} of dequeue when there are pending nodes. We also need to update the rules in figure.}

\begin{wrapfigure}{l}{7cm}
\vspace{-8mm}
\includegraphics[width=7cm]{fig-queue12.pdf}
%
%\vspace{2mm}
%\includegraphics[width=7cm]{fig-queue2.pdf}
\vspace{-5mm}
\caption{Simulating queue histories with $AbsQ$.}
\label{fig:queueSim}
\vspace{-10mm}
\end{wrapfigure}
Figure~\ref{fig:queueSim} pictures two executions of $AbsQ$ for two extended histories (that include dequeue linearization points). The state of $AbsQ$ after each action is pictured as a graph below the action. The nodes of this graph represent enqueue operations and the edges happens-before constraints. Each node is labeled by a value (the input of the enqueue) and a flag {\tt PEND} or {\tt COMP} showing whether the operation is pending or completed. For instance, in the case of the first history, the dequeue linearization point $lin(deq,y,3)$ is enabled because the current happens-before contains a \emph{minimal} enqueue operation with input $y$. Note that a linearization point $lin(deq,x,3)$ is also enabled at this state.

@@

We show that $\mathit{fs}$ is indeed a $C\cup R\cup Lin(deq)$-forward simulation. Let $s$ and $t$ be states of $\mathit{HWQ}$ and $AbsQ$, respectively, such that $(s,t)\in\mathit{fs}$. 
We omit discussing the trivial case of transitions labeled by call and return actions which are simulated by similar transitions of $AbsQ$ (for the return a dequeue operation $k$, we use the equality between the local variable ${\tt x}(k)$ in $s$ and the component $rv(k)$ in $t$). 
%\textcolor{red}{ I think it is good to mention again that call/return actions in HWQ correspond to the same call/return actions in AbsQ (without any other internal action). I also think that invoke enqueue operation is non-trivial. Preservation of the strict partial order and all of the above items a, b and c needs to be rechecked.}

### 2017-01-19 draftCAV/stack.tex
small modifs
started stacks

We define an abstract stack implementation $AbsS$ over alphabet $C\cup R\cup Com(pop)$ that essentially, similarly to $AbsQ$, maintains the happens-before order of the push operations whose value has not been yet removed. However, since the commit points are not necessarily linearization points, the pop operations are treated differently. Intuitively, each pop operation starts by taking a snapshot of the completed push operations which are maximal in the happens-before order and continuously tracks the push operations which are overlapping with it. The commit point is enabled only if the return value corresponds to one of the push operations in the initial snapshot, or to a push happening earlier when values from the initial snapshot are removed by other pop operations, or to one of the push operations that overlaps with the pop.

@@

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp) \\ 
    }{
      O,<,\ell,rv,cp,be,ov 
      \xrightarrow{inv(push,d,k)} \\
      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1],be,ov[\forall k'.\ k'\mapsto ov(k')\cup \{k\}]
    }\hspace{5mm}

@@

    \inferrule[lin-pop1]{
       cp(k) = 1 \\
       d\neq{\tt EMPTY} \\
       k'\in be(k)\cup ov(k)\land \ell_1(k')=d
    }{
      O,<,\ell,rv,cp,be,ov
      \xrightarrow{com(pop,d,k)} \\
      O\setminus \{k'\},<\uparrow k',\ell,rv[k\mapsto d],cp[k\mapsto 2],be[\forall k''.\ k'\in be(k'') \Rightarrow k'' \mapsto (be(k'')\cup pred_{<}(k'))\setminus\{k'\}],ov
    }\hspace{5mm}

@@

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The transition relation of $AbsQ$. We use the following notions: $f[\forall k.guard(k)=> k\mapsto expr(k)]$ is the function $g$ defined by $g(k)=f(k)$ for all $k$ such that $guard(k)$ is false, and $g(k)=expr(k)$ otherwise, $maxCo(O)$ is the set of maximally completed operations in $O$, i.e., $maxCo(O)=\set{k\in O: \ell_1(k)={\tt COMP}, \forall k'\in O.\ k' < k \vee \ell_1(k')={\tt PEND}}$, ${\tt PEND}(O)=\{k\in O: \ell_2(k)={\tt PEND}\}$, and $pred_{<}(k')$ is the set of immediate predecessors of $k'$ according to $<$, i.e., $pred_{<}(k')=\set{k\in O: k < k'\land \forall k''\in O.\ k'' > k' \vee k'' < k}$.\textcolor{red}{call-push should have $d \neq EMPTY$ in the premise. If I understand correctly, there is a problem change in $be$ in lin-pop1 rule. What I understand is we move $k''$ to every predecesoor of $k'$. But we should not move it to predecessor of $k'$ if this predecessor has a successor $m$ such that $m \in be(k'')$. Is this case reflected in your definition. Please read sub-bullet 4 of bullet 4 in the definition of $L_I$ in my document. Lin-pop2 seems ok for me. Last note: Bibliography seems not correct to me.} }
  \label{fig:transitions:AbsS}
\vspace{-6mm}
\end{figure}

>>>>

We define an abstract stack implementation $AbsS$ over alphabet $C\cup R\cup Com(pop)$ that essentially, similarly to $AbsQ$, maintains the happens-before order of the push operations whose value has not been yet removed. However, since the commit points are not necessarily linearization points, the pop operations are treated differently. Intuitively, each pop operation starts by taking a snapshot of the completed push operations which are maximal in the happens-before order and continuously tracks the push operations which are overlapping with it. The commit point is enabled only if the return value corresponds to one of the push operations in the initial snapshot, or to a push happening earlier when the values from the initial snapshot are removed by other pop operations, or to one of the push operations that overlaps with the pop. TODO EMPTY

\begin{wrapfigure}{l}{7cm}
\vspace{-8mm}
\includegraphics[width=7cm]{fig-stack.pdf}
%
%\vspace{2mm}
%\includegraphics[width=7cm]{fig-queue2.pdf}
\vspace{-5mm}
\caption{Simulating stack histories with $AbsS$.}
\label{fig:stackSim}
\vspace{-6mm}
\end{wrapfigure}
Figure~\ref{fig:stackSim} pictures two executions of $AbsQ$ for two extended histories (that include dequeue linearization points). The state of $AbsQ$ after each action is pictured as a graph below the action. The nodes of this graph represent enqueue operations and the edges happens-before constraints. Each node is labeled by a value (the input of the enqueue) and a flag {\tt PEND} or {\tt COMP} showing whether the operation is pending or completed. For instance, in the case of the first history, the dequeue linearization point $lin(deq,y,3)$ is enabled because the current happens-before contains a \emph{minimal} enqueue operation with input $y$. Note that a linearization point $lin(deq,x,3)$ is also enabled at this state.

@@

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp) \\ 
      d\neq {\tt EMPTY}
    }{
      O,<,\ell,rv,cp,be,ov 
      \xrightarrow{inv(push,d,k)} \\
      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1],be,ov[\forall k'.\ k'\mapsto ov(k')\cup \{k\}]
    }\hspace{5mm}

@@

    \inferrule[lin-pop1]{
       cp(k) = 1 \\
       d\neq{\tt EMPTY} \\
       k'\in be(k)\cup ov(k)\land \ell_1(k')=d \\
       \forall k_1.\ k'\not\in be(k_1) \Rightarrow be'(k_1)=be(k_1) \\     \\
       \forall k_1.\ k'\in be(k_1) \Rightarrow be'(k_1)=(be(k_1)\setminus\{k'\})\cup \{k_2: k_2\in pred_{<}(k') \land \forall k_3. k_2<k_3 => k_3\not\in be(k_1)\} \\
    }{
      O,<,\ell,rv,cp,be,ov
      \xrightarrow{com(pop,d,k)} \\
      O\setminus \{k'\},<\uparrow k',\ell,rv[k\mapsto d],cp[k\mapsto 2],be',ov %be[\forall k''.\ k'\in be(k'') \Rightarrow k'' \mapsto (be(k'')\cup pred_{<}(k'))\setminus\{k'\}]
    }\hspace{5mm}

@@

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The transition relation of $AbsQ$. We use the following notions: $f[\forall k.guard(k)=> k\mapsto expr(k)]$ is the function $g$ defined by $g(k)=f(k)$ for all $k$ such that $guard(k)$ is false, and $g(k)=expr(k)$ otherwise, $maxCo(O)$ is the set of maximally completed operations in $O$, i.e., $maxCo(O)=\set{k\in O: \ell_1(k)={\tt COMP}, \forall k'\in O.\ k' < k \vee \ell_1(k')={\tt PEND}}$, ${\tt PEND}(O)=\{k\in O: \ell_2(k)={\tt PEND}\}$, and $pred_{<}(k')$ is the set of immediate predecessors of $k'$ according to $<$, i.e., $pred_{<}(k')=\set{k\in O: k < k'\land \forall k''\in O.\ k'' > k' \vee k'' < k}$.%\textcolor{red}{call-push should have $d \neq EMPTY$ in the premise. If I understand correctly, there is a problem change in $be$ in lin-pop1 rule. What I understand is we move $k''$ to every predecesoor of $k'$. But we should not move it to predecessor of $k'$ if this predecessor has a successor $m$ such that $m \in be(k'')$. Is this case reflected in your definition. Please read sub-bullet 4 of bullet 4 in the definition of $L_I$ in my document. Lin-pop2 seems ok for me. Last note: Bibliography seems not correct to me.} 
  }
  \label{fig:transitions:AbsS}
\vspace{-6mm}
\end{figure}
